---
title: "Stats 107 Project"
author: "Omar Sepulveda, Riccardo Gutierrez"
date: "2025-11-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
###ABSTRACT 
The NCAA SEC is one of the premier conferences in college football from recent years, but how much is each program represented in the draft? What we are looking to accomplish is to highlight which schools in the NCAA SEC (South Eastern Conference) have gained representation in the draft and which programs have lost representation in the draft over the last 15 years. We collected our data from Pro-Football-Reference.com and combined each program in the SEC data sets into a singular Excel file and analyzed the data by distinhuishing which players came from the which college program, seeing if the player was drafted, then drafted within the years 2011-2025.

```{r}
library(readxl)
library(dplyr)

# Read the Excel file
data <- read_excel("College Dataset.xlsx")

# Rename columns for clarity (based on the header)
colnames(data) <- c("Pos", "Player", "College_Stats", "AP1", "PB", "St", "wAV", "Ht", "Wt", "Yrs_Num", "Years_Range", "School")

# Extract the start year from Years_Range
data <- data %>%
  mutate(start_year = as.integer(sub("-.*", "", Years_Range)))

# Filter to keep players who started in 2011 or later
filtered_data <- data %>%
  filter(start_year >= 2011)

# Select only the specified columns
fixed_data <- filtered_data %>%
  select(Pos, Player, Yrs_Num, Years_Range, School)

# Rename columns as per user labels if desired
colnames(fixed_data) <- c("Position", "Player", "Years Played", "Years Played (From-To)", "School")

# View the selected data (or you can use write_excel_csv to save it)
head(fixed_data)
```

# Data

The data set being used for this analysis is "College_Dataset.xlsx", which contains player names and the colleges they attended. This data was retrieved from the Pro-Football-Reference website that collect's and releases all NFL data including players and what schools they attended before entering the NFL. The variables that are relevant in the data set are as follow, the players name, the school they came from, and the years the player started to play in the NFL to the last year they played in the NFL. As far as cleaning the data, what we did was get rid of unnecessary variables such as AP1, PB, St, wAV, Ht, and Wt, and only kept the relevant ones as stated before. Then since the data set has players from the 20th century, and 2000 to 2010, what we did was just extracted the starting year from each player’s career range, and filtered the data to include only players who began their NFL careers in 2011 or later. No, the data will not be generated through a randomized simulation.
