---
title: "Trends in NFL Player Representation from SEC Schools (2011–2025)"
author: "Omar Sepulveda, Riccardo Gutierrez, Naomi Menard"
date: "2025-12-05"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, warning = FALSE, echo = FALSE}
# This chunk loads the script to attach necessary packages like tidyverse and gridExtra. 
# It sets up the environment for data manipulation and visualization.
# Outcome: Packages are installed and enabled

source("00_requirements.R")
```

# Abstract

The NCAA SEC is one of the premier conferences in college football from recent years, but how much is each program represented in the draft? Our data was collected from Pro-Football-Reference.com, which is a database that tracks every NFL player's stats, awards, drafting, and teams during college and NFL. Our analysis examines the number of draftees from each SEC football program for every individual year from 2011 through 2025 and compares trends within the graphs to further analyze the popularity and production of each program (in terms of outputting NFL talent). The results yield two common patterns in the data: either the program shows a quadratic trend that peaks in the late 2010s, or an exponential trend in which draft representation steadily grows from 2011 onward.

# Introduction

The NFL is the highest viewed sport in the U.S. as it has been nicknamed Americas sport by the media. In turn college football also exudes major viewer retention and support since the players in college are the future of the NFL, also college students love to support their school. Over the last 15 years there has been one NCAA football conference that has prominently outputted the most NFL talent, that being the South Eastern Conference (SEC). Numerous factors apply when it comes to recruitment, such as head coach prestige, academic prestige, and program prestige, all these factors apply but the most important factor is players drafted. Powerhouses like Alabama and Georgia have remained constant machines when fostering NFL talent, but over recent years LSU and Texas have provided incredible NFL talent. From 2011 to the present year 2025 the SEC has remained a constant pool of NFL talent as they are the programs that every high school football player strives to be apart and compete with the talent within the SEC. The trends for each SEC program in terms of draft representation will differ considering that each program competes against all the other programs within the SEC, therefore more recruits will strive to attend the programs that are winning and the programs that create the most NFL talent. Therefore, if we can analyze the trends for each SEC program in terms of draft representation we can comprehend how each program performed during the period 2011-2025 and which programs grew or withered in draft representation.

# Data
```{r message = FALSE, warning = FALSE, echo = FALSE}
# This chunk sources the data cleaning script to process the raw dataset. 
# It takes raw data, keeps/renames columns, filters to 2011+, and saves the cleaned data.
# Outcome: A cleaned dataset is made and saved as "cleaned_data.RDS" for later use.

source("01_funct_DataCleaning.R")
```
The data set used for this analysis, “College_Dataset.xlsx,” contains 1470 observations of information on NFL players and the colleges they attended, with players who began their NFL careers from 2011 to 2025, spanning 14 years. The data was obtained from Pro-Football-Reference, a reliable source that collects and publishes official NFL statistics, including player backgrounds and school affiliations. Relevant variables include each player’s name, the college they attended, and the range of years they played in the NFL. During the cleaning process, unnecessary columns such as AP1, PB, St, wAV, Ht, and Wt were removed, keeping only the relevant variables, while still keeping the 1470 observations in tact. Additionally, since the data set included players from earlier eras, we extracted the starting year from each player’s career range and filtered the data to include only those who began their NFL careers in 2011 or later. The observations are not evenly distributed across the years, with the number of new players varying annually based on factors such as undrafted signings. The data includes players from 16 unique schools. The data provides the individual starting year for each player, as the yearly data enables more granular and insightful analysis. The data is entirely real and will not be generated through any randomized simulation.

# Visualization & Analysis


```{r, echo = FALSE}
source("02_funct_Plots.R")
```

Here we have displayed individual bar charts for each SEC school, showing the number of players that were selected in the NFL Draft in each individual year from 2011 through 2025. Based on first impressions, without considering the exact numeric counts, there are clear trends indicating which programs increased or decreased in draft representation over time. Georgia, Florida, Alabama, South Carolina, Ole Miss, and Texas have shown significant growth in draft representation, while many of the other SEC programs have displayed either steady or declining results across the years.

```{r echo = FALSE}
# This chunk sources the table script to summarize data numerically. 
# It creates a tibble table with player counts by school and year 
# Outcome: A table is generated and printed, providing exact counts for comparison.
source("03_numeric_Table.R")
```

```{r echo= FALSE}
source("04_percentage_changes.R")
```

```{r echo= FALSE}
source("05_percent_map.R")
```
The heat map displays the year-to-year percentage change in NFL draft selections for each SEC program from 2011 to 2025. Each tile represents how a school’s draft output changed compared to the previous year. Blue shades indicate positive growth in drafted players, red shades indicate a decline, and white represents little or no change. This visualization makes it easy to compare trends both across time and between schools, revealing programs with consistent improvement, sudden drops, or highly variable draft production. While some individual year-to-year jumps may appear large,especially when starting from a low number of drafted players,the overall long-term trend is more important. For example, Ole Miss shows very few red areas and more blue or white regions, suggesting relatively steady or improving draft representation over the 15-year period. It is important to note that the concentration or intensity of colors is less significant than identifying which programs display the least amount of red, indicating greater consistency or growth.



# Analysis

For the analysis, we plan to use descriptive statistics and visual models to track changes in the number of NFL players from each SEC school over the three time periods. We'll look at the overall trend for each school using a simple linear regression, which will help show whether the number of NFL players from that school is going up or down over time. We'll also use percentage change calculations to measure growth rates and rank the schools based on overall gains or declines. These methods should make it clear which schools have built a stronger NFL presence over time. We expect that powerhouse programs like Alabama and Georgia will show a steady growth, while schools like Vanderbilt and Auburn may show slight growth or even a decline. For further analysis discussing the histograms hopefully the year by year displays for each collegiate program in the conference will further contribute to our theory of steady growth for the elite programs while the lesser known programs won't display such trends. We expect to find that most if not all programs within the conference will display some form of growth within the 2011-2025 time period and then a variation of growth and decay during the covid year as not all programs held the same appeal during the pandemic. With big name schools such as Alabama, Georgia and Texas will display steady growth and the values will remain relatively high compared to lesser known schools such as Auburn and Vanderbilt.

# Contributions

Omar Sepulveda: Gathered data sets, completed data section, code used to clean data set, code for the visuals, completed analysis section, and wrote step by step process in README.md

Riccardo Gutierrez: Gathered data sets, completed abstract section, and contributed to visualization and analysis section.

Naomi Menard: Completed introduction, gathered data sets, contributed to data, visuals/code, analysis sections.

```{r}
names(data)
```

```{r}
source("06_poisson.R")
```