---
title: "Trends in NFL Player Representation from SEC Schools (2011–2025)"
author: "Omar Sepulveda, Riccardo Gutierrez, Naomi Menard"
date: "2025-11-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("00_requirements.R")
```

# Abstract

The NCAA SEC is one of the premier conferences in college football from recent years, but how much is each program represented in the draft? We collected our data from Pro-Football-Reference.com and combined each program in the SEC data sets into a singular Excel file and analyzed the data by distinguishing which players came from which college program, seeing if the player was drafted, then drafted within the years 2011-2025. 

# Introduction

The purpose of this analysis is to highlight which schools in the NCAA SEC (South Eastern Conference) have gained representation in the draft and which programs have lost representation in the draft over the last 15 years. This will answer the question: "How has the number of NFL players from each SEC school changed over time (e.g., 2011–2015, 2016–2020, and 2021–2025), and which schools have shown the most growth or decline in player representation?" This analysis may benefit various groups, including NFL scouts/teams, sports analysts/journalists, or other researchers looking to see which college improved the most. The goal is to understand patterns and changes throughout the 2011 - 2015, 2016-2020, and 2021-2025 time periods.

# Data
```{r}
source("01_funct_DataCleaning.R")
```
The data set used for this analysis, “College_Dataset.xlsx,” contains information on NFL players and the colleges they attended. The data was obtained from Pro-Football-Reference, a reliable source that collects and publishes official NFL statistics, including player backgrounds and school affiliations. Relevant variables include each player’s name, the college they attended, and the range of years they played in the NFL. During the cleaning process, unnecessary columns such as AP1, PB, St, wAV, Ht, and Wt were removed, keeping only the relevant variables. Additionally, since the data set included players from earlier eras, we extracted the starting year from each player’s career range and filtered the data to include only those who began their NFL careers in 2011 or later. The data is entirely real and will not be generated through any randomized simulation.

# Visualization & Analysis

```{r}
source("02_funct_Plots.R")
```

```{r}
source("03_numeric_Table.R")
```

These preliminary visualizations support our future analysis by allowing us to compare the number of NFL players produced by various SEC colleges across different time periods. By examining the bar plots, we can identify early patterns of growth or decline in player representation for each school. In later stages of the analysis, we will compare these bar plots in greater detail to draw conclusions about which colleges have shown the most progress or decline in producing NFL players over the years.

# Analysis

For the analysis, we plan to use descriptive statistics and visual models to track changes in the number of NFL players from each SEC school over the three time periods. We'll look at the overall trend for each school using a simple linear regression, which will help show whether the number of NFL players from that school is going up or down over time. We'll also use percentage change calculations to measure growth rates and rank the schools based on overall gains or declines. These methods should make it clear which schools have built a stronger NFL presence over time. We expect that powerhouse programs like Alabama and Georgia will show a steady growth, while schools like Vanderbilt and Auburn may show slight growth or even a decline.

# Contributions

Omar Sepulveda: Gathered data sets, completed data section, code used to clean data set, code for the visuals, and completed analysis section


