---
title: "Trends in NFL Player Representation from SEC Schools (2011–2025)"
author: "Omar Sepulveda, Riccardo Gutierrez, Naomi Menard"
date: "2025-12-05"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, warning = FALSE, echo = FALSE}
# This chunk loads the script to attach necessary packages like tidyverse and gridExtra. 
# It sets up the environment for data manipulation and visualization.
# Outcome: Packages are installed and enabled

source("00_requirements.R")
```

# Abstract

The NCAA SEC is one of the premier conferences in college football from recent years, but how much is each program represented in the draft? Our data was collected from Pro-Football-Reference.com, which is a database that tracks every NFL player's stats, awards, drafting, and teams during college and NFL. Our analysis examines the number of draftees from each SEC football program for every individual year from 2011 through 2025 and compares trends within the graphs to further analyze the popularity and production of each program (in terms of outputting NFL talent). The results yield two common patterns in the data: either the program shows a quadratic trend that peaks in the late 2010s, or an exponential trend in which draft representation steadily grows from 2011 onward.

# Introduction

The NFL is the highest viewed sport in the U.S. as it has been nicknamed Americas sport by the media. In turn college football also exudes major viewer retention and support since the players in college are the future of the NFL, also college students love to support their school. Over the last 15 years there has been one NCAA football conference that has prominently outputted the most NFL talent, that being the South Eastern Conference (SEC). Numerous factors apply when it comes to recruitment, such as head coach prestige, academic prestige, and program prestige, all these factors apply but the most important factor is players drafted. Powerhouses like Alabama and Georgia have remained constant machines when fostering NFL talent, but over recent years LSU and Texas have provided incredible NFL talent. From 2011 to the present year 2025 the SEC has remained a constant pool of NFL talent as they are the programs that every high school football player strives to be apart and compete with the talent within the SEC. The trends for each SEC program in terms of draft representation will differ considering that each program competes against all the other programs within the SEC, therefore more recruits will strive to attend the programs that are winning and the programs that create the most NFL talent. Therefore, if we can analyze the trends for each SEC program in terms of draft representation we can comprehend how each program performed during the period 2011-2025 and which programs grew or withered in draft representation.

# Data
```{r message = FALSE, warning = FALSE, echo = FALSE}
source("01_funct_DataCleaning.R")
```
The data set used for this analysis, “College_Dataset.xlsx,” contains 1470 observations of information on NFL players and the colleges they attended, with players who began their NFL careers from 2011 to 2025, spanning 14 years. The data was obtained from Pro-Football-Reference, a reliable source that collects and publishes official NFL statistics, including player backgrounds and school affiliations. Relevant variables include each player’s name, the college they attended, and the range of years they played in the NFL. During the cleaning process, unnecessary columns such as AP1, PB, St, wAV, Ht, and Wt were removed, keeping only the relevant variables, while still keeping the 1470 observations in tact. Additionally, since the data set included players from earlier eras, we extracted the starting year from each player’s career range and filtered the data to include only those who began their NFL careers in 2011 or later. The observations are not evenly distributed across the years, with the number of new players varying annually based on factors such as undrafted signings. The data includes players from 16 unique schools. The data provides the individual starting year for each player, as the yearly data enables more granular and insightful analysis. The data is entirely real and will not be generated through any randomized simulation.

# Visualization & Analysis


```{r, echo = FALSE}
source("02_funct_Plots.R")
```

Figure 1. Bar chart representation of the number of players drafted per year from 2011-2025. Each graph is incremented by 5, however trends are prevalent in each graph representing the SEC programs.

Here we have displayed individual bar charts for each SEC school, showing the number of players that were selected in the NFL Draft in each individual year from 2011 through 2025. Based on first impressions, without considering the exact numeric counts, there are clear trends indicating which programs increased or decreased in draft representation over time. Georgia, Florida, Alabama, South Carolina, Ole Miss, and Texas have shown significant growth in draft representation, while many of the other SEC programs have displayed either steady or declining results across the years.

```{r echo = FALSE}
# This chunk sources the table script to summarize data numerically. 
# It creates a tibble table with player counts by school and year 
# Outcome: A table is generated and printed, providing exact counts for comparison.
source("03_numeric_Table.R")
```
Figure 2. A numeric table that presents the exact number of draftees per each program for every year from 2011 to 2025.

The numeric table provides a detailed yearly breakdown of the number of drafted players from each SEC school between 2011 and 2025. Each row represents a school, while the "Start" column corresponds to a specific draft year. The "Players" values indicate exactly how many players that program contributed to the NFL Draft in that given year. Years with no drafted players are represented as zeros, ensuring a complete and consistent comparison across all schools. This table presents the raw counts directly, making it useful for understanding absolute draft contributions and for comparing overall volume across programs over the entire 15-year span. 

```{r echo= FALSE}
source("04_percentage_changes.R")
```

```{r echo= FALSE }
source("05_percent_map.R")
```
Figure 3. A heat map that covers the year by year percentage change for each SEC program with the darker the color of each box the greater the percentage change present. There are some boxes that are presented as white because there were no draftees during that year, therefore the box is white.

The heat map displays the year-to-year percentage change in NFL draft selections for each SEC program from 2011 to 2025. Each tile represents how a school’s draft output changed compared to the previous year. Blue shades indicate positive growth in drafted players, red shades indicate a decline, and white represents little or no change. This visualization makes it easy to compare trends both across time and between schools, revealing programs with consistent improvement, sudden drops, or highly variable draft production. While some individual year-to-year jumps may appear large,especially when starting from a low number of drafted players,the overall long-term trend is more important. For example, Ole Miss shows very few red areas and more blue or white regions, suggesting relatively steady or improving draft representation over the 15-year period. It is important to note that the concentration or intensity of colors is less significant than identifying which programs display the least amount of red, indicating greater consistency or growth.

```{r echo= FALSE}
source("06_poisson.R")
```

```{r echo =FALSE, fig.width=2, fig.height=2}
source("07_poisson_visual.R")
```
Figure 4. A Poisson regression graph that highlights the trend of the number of draftees within each SEC program over the given interval from 2011-2025. Some graphs depict an exponential curve while others depict a more linear curve considering the change in draftees is les drastic compared to others.

These Poisson regression graphs are a greater visual representation of the trends highlighted within the bar graphs from Figure 1. as we can visually see the curvature within each SEC programs change in number of draftees. Teams like Alabama, Georgia, Kentucky, Ole Miss, Tennessee, and Texas are visualized with a more exponential trend because throughout the years these are the teams that have created a winning culture throughout the years, therefore more players are going to these schools and therefore more NFL draft talent is being produced from these programs. Other schools like Arkansas, Auburn, Florida, LSU, Mississippi St., Missouri, Oklahoma, South Carolina, Texas A&M, Vanderbilt all show a more liner trend because even though they aren't the programs that are winning the most, they are in the conference where the most popular, successful teams are competing, therefore they are garnering more exposure to NFL scouts so that they have greater chances to be drafted to an NFL team. However, Arkansas, Auburn, and Florida are all experiencing negative trends, compared to all other programs experiencing positive trends, these are the teams that have suffered the most from the success of the SEC as these teams have produced NFL talent, but there programs have struggled to win and be competitive in the conference leading to less recruitment and therefore less NFL talent. The exponential trends within these graphs highlight a drastic increase in recruitment and NFL draft talent production from schools experiencing this type of trend, inferring that these teams have been the most successful and have seen an increase in NFL draft representation. While teams experiencing more linear trends are inferred to have experience a more steady increase or decrease in NFL draft representation due to either an increase of recruitment and exposure or a decrease in recruitment and exposure. 

# Conclusion

Based on the analysis and visualization conducted, it appears that most of the programs presented within the South Eastern Conference (SEC) did experience some form of positive trend, the only outliers being Arkansas, Auburn, and Florida. This notion is supported by the Poisson regression graphs that visualize either some form of exponentially positive trend or some form of linear, positive trend that highlights the increase of NFL draft representation for 13 out of the 16 programs within the SEC. This suggests that the SEC and the winning, competitive culture surrounding the conference has caused some form of increase in NFL draft representation for most of the programs competing within the SEC.

To further analyze the results beyond the trends Figures 1. and 3. highlight the exact jumps in NFL draft representation year by year. The bar graphs and the table highlight the exact jumps in representation on a year by year scale. While the heat map depicts the percentage change in NFL draft representation, not contingent on whether the change is positive or negative, the change just represents the aggregate jump in draft representation. 

In conclusion, based on the analysis presented within the graphs, maps, and tables we can assume that Georgia has experienced the greatest growth in NFL draft representation during the time period 2011-2025. On the other hand Arkansas is the program that has withered the most in NFL draft representation during the time period 2011-2025. To further delve into the data, future analysis can create poisson regression models to obtain the exact numbers that relate to the Poisson regression graphs highlighting the regression lines. Consequently this analysis is not perfect, some further analysis could consider the outlying factors that go into college recruitment and therefore exposure to NFL scouts, such as the history of the programs, along with the NFL talent produced from the programs. Even factors relating to the college or university itself, such as campus culture and academic prestige can all contribute to why a college recruit would choose to committ or transfer to a program within the notorious, competitive, and successful NCAA South Eastern Conference.

# Contribution

Omar Sepulveda: Gathered data sets, completed data section, code used to clean data set, code for the visuals, completed analysis section, and wrote step by step process in README.md

Riccardo Gutierrez: Gathered data sets, completed abstract section, visuals code, contributed to visualization and analysis section, completed Conclusion.

Naomi Menard: Completed introduction, gathered data sets, contributed to data, visuals/code, analysis sections. Created descriptions for visuals.